Scaling up a GF grammar and preparing it for cool stuff for Swedish

### 2011-11-28 doc/report/report.tex
backup of docs, not newer than showup-branch

\documentclass{report} 
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage[margin=1.5in]{geometry}
\usepackage{lingmacros}
\usepackage{subfig}

@@

@@

GF is developed with multilinguality in mind and makes an important distiction between
\textit{abstract} and \textit{concrete} syntaxes. The abstract is meant to be
shared by many languages, it models the semantics of the domain without blanda in
language specific features such as agreement or word order.

\begin{figure}[h]
\begin{verbatim}
 abstract TestGrammar {
  cat N ; V ; S ;

  fun 
    Pred : N -> V -> S ;
 }
\end{verbatim}
\caption{A small example of an abstract syntax}
\end{figure}

The example shows an abstract grammar defining three \textit{Categories},
one for nouns, one for verbs and one for sentences. An abstract grammar also gives
the types of functions. In this case we have \verb|Pred| which
tells us that by taking a noun and a verb we can form a sentence. It does not
give any information of how this is done, or how any of the categories should look.
maybe continue example with swedish and english, have DetNP : N -> N wich
gives 'the cat', 'katten'. Thereby introduce tables, parameters, fields.
Show abstract tree.

We now have two implementations of the abstract. The resulting GF grammar is able to
both parse a string to the abstract tree and to go in the other direction; to produce
a string of natural language given a abstract tree. This step is called linearization.
Translation is a consequence of this. Picture of sw. string -> tree -> eng. tree.

@@

A GF grammar usually covers a very restricted set of language, where few ambiguities
occur and much care can be taken to preserving the semantics between the different languages.
By using a controlled language like this,  good and reliable translation can be
ensured. Some examples of projects using GF are the math .. \cite{olga}, the
Online Syllogism Solver\cite{malin!}.  The framework is also used in the
european project MOLTO for online translation between upto 15 language.\\

One of the best... features of the framework is the resource libraries. Provided in the 
GF package, those gives a good start for all who want to write an application grammar.
Instead of letting the users reimplementing the morphological and the fundamental
syntactic structures of a language for each application, all this can be found
as libraries. So far, 20 languages are implemented all sharing the same abstract syntax.
They describe how to conjugate words, and how to construct phrases and sentences.
There are also a small test lexicon included, containing some hundered of common words.\\

@@

\subsection{Swedish}
Swedish is a North-Germanic language spoken,
closely related to Norweigan and Danish. The languages share most of their
grammatical structures and are mutual intelligibility. Swedish is also one
a official language in Finland and approximately 10 million people speak Swedish.
Charasterictic for Swedish is the fact that it is a verb-second language: the
second constituent of a declarative main clause must consist of a verb.
The first constituent of the clause is usually made up of the subject,
although it likewise could consist of adverbial phrases or objects that are
being empathazied. \\

\enumsentence{
\begin{tabular}{lll}
  Johan & gick & sakta på gatan.\\
\emph{Johan} & \emph{walked} & \emph{slowely on the street.}\\
  Sakta & gick & Johan på gatan.\\
\emph{Slowly} & \emph{walked} & \emph{Johan on the street.}\\
  På gatan & gick & Johan sakta.\\
\emph{On the street} & \emph{walked}  & \emph{Johan slowly.}\\
\end{tabular}}

Inverted word order marks questions
\enumsentence{Gick Johan på gatan?\\
        \emph{Did Johan walk on the street?}}
The word order in subordinate clauses in also slightly different,
\enumsentence{Jag såg Anna. Hon hälsade \textbf{inte}.\\
        \emph{I saw Anna. She did not say hello.}\\
              Jag såg att Anna \textbf{inte} hälsade.\\
        \emph{I saw that Anna did not say hello.}}

Some other features typical for Swedish is it's use of passive, especially the
\textit{s-passive}, which is formed by adding an \emph{s} at the end of a verb
\enumsentence{Han jagade\textbf{s} av ett lejon.\\
        \emph{He was hunted by a lion.}}
Jfr.
\enumsentence{Ett lejon jagade honom.\\
        \emph{A lion was hunting him.}}

Constructions with 'det är/var' ('it is/was') is very common\cite{H&H309d}:
\enumsentence{Det var roligt att höra.\\
        \emph{I'm glad to hear that.}}
'Det' \emph{it} is also used as formal subject for intransitive verbs,
with the real subject put in the position of an object.
\enumsentence{Det står en älg på fältet.\\
        \emph{There is a mouse on the field.}}

The Scandinavian language have reflexive pronouns\cite{H&H310}
and special reflexive forms of the possessive pronouns for the 3rd
person\cite{H&H319}. Those are distinct from the normal 3rd person forms.
\enumsentence{
\begin{tabular}{ll}
Han slog sig. & Han såg sitt barn.\\
\vspace{3mm}
\emph{He hit him self.} & \emph{He saw his (own) child.}\\
Han slog honom. & Han såg hans barn.\\
\emph{He hit him (another person).}&\emph{He saw his (another person's) child.}
\end{tabular}}
The 1st and 2nd persons use the normal personal pronoun in object form.

@@

\subsection{Saldo}
A good parser needs a good lexicon. We have used of Saldo\cite{saldo}, a
large electronic lexicon developed and maintained at Gothenburg University. It is
built upon Svenskt Associations Lexicon and contains information about modern
Swedish and the words are equipped with both sematic and a morphological
information. The user can find compounding analyses, examples of usage in corpora,
and graphs of sematically connected words.\\
Example of katt? \\
For our purpose however, only the morphological data was needed.\\
Example of morpho for katt.\\
Saldo provides this in XML format under LPGL. The data can be processed and translated to GF format as described in 
section \ref{sec:prog.saldo}.
Other usages?

@@

\subsection{GF for Swedish}
The GF resource grammars contains a fundamental description of Swedish,
covering the morphology, many phrases and clauses.
Covers sentences like
han har inte ätit två gula äpplen idag
word order, agreement, tempus, adverbs, adjectivs...
and relatives
han såg katten han tyckte om
Also info in the Extra module, where more language specific constructions
can be found.
honom ville hon inte tänka på.

@@

For verbs, but also others there is an important distinction for valency.
Either directly to VP if V, or needs to be given arguments. Can then add adverbs.
Maybe talk about ställa.

@@

%%PASSIV
Is passive also more used in Swedish?
There are two ways of forming passive verb phrases in Swedish: the 
periphrastic passive, formed by using a modale auxiliary verb 
\enumsentence{Bilden blev tagen av  av rymdsonden Galileo}
\enumsentence{The picture was taken by spacecraft Galileo.}
or by adding an \emph{s} to
the verb:
\enumsentence{Bilden togs av rymdsonden Galileo}
\enumsentence{The picture [took+s] by spacecraft Galileo.}
This second variant, the s-passive, is commonly used in Swedish, some research
suggest that it is used in more than 80\% of the times \cite{laanemets}.
The other scandinavian languages prefers the periphrastic passive, especially in spoken langauge.
\enumsentence{Bildet ble tatt av romsonden (NO)}
The resource grammar for Scandinavian therefore allowed/expressed constructions using auxiliary verb. \\
\verb|PassV2 : V2  -> VP | \\
\verb|         äta -> blev äten| \\
This rules allows two-place verbs to be used in passive by using \emph{bli} (\emph{become}), and thereby
turned into comlpete verb phrases; they no longer need an object.
The s-passive has now been added as the standard case for the Swedish GF
grammar. Periphrastics passive is still allowed, but as an alternative rather
than the default.
The grammar also allows using three-place verbs \verb|V3| , as \emph{give} to
form passives:
\enumsentence{Betyget 3 ges till många elever}
\label{ex:passV3}
by letting the first object be `removed' and given by the subject. 
Sentences where the second object is put first in the sentence, like 
\enumsentence{Till många elever ges betyget 3}
is a variant of example \ref{ex:passV3}, \emph{`många elever'} is still the object but put in focus. The grammar
parse this by applying \verb|FocObj|.
%%

@@

Since those constructions are inspired by other languages, some of them may sound quite
non-natural in Swedish, such as the \verb|RelS|. This combines a sentence \emph{`Hon sover'}
and a relative sentence \emph{`som/vilket är bra'} which wrongly expressed 
\emph{`She sleeps, which is good'} as
\emph{`Hon sover som är bra'}. The error comes from the fact that \emph{`som'} is
embedded in the relative clause. To fix this, an extra parameter was needed to tell whether
\emph{`som'} or \emph{`vilket'} should be used. \\
Another complication stems from \verb|RelCl| which express the english \emph{`such that'}.
The Swedish version \emph{`sådan att'} sounds stelbent, when used outside of logic books.
\enumsentence{Jag vill ha en katt sådan att den inte fäller.} 
An more natural sounding alternative would be
\enumsentence{Jag vill ha en sådan katt som inte fäller.}
and hopefully this will be implemented soon, otherwise: why is it hard and how we can say
'en katt sådan att det regnar'. It is relative but has on empty field, still congruation is
needed.\\

Formal subjects \cite{SAG-19,H&H-309d} is often used in Swedish
\enumsentence{Det sitter en fågel på taket}
\emph{`There is a bird sitting on the roof'}
or, literally
\emph{`It sits a bird on the roof'}.\\
\emph{`Det'} has the position of the subject, and the real subject, 
\emph{`en fågel'} the one of an object.
Transitive verbs may not be used like this
\enumsentence{*Det äter en fågel på taket.}
unless their in passive form
\enumsentence{Det äts en fågel på taket.}
%det finns/saknas/fattas
There are also restrictions on the nounphrase of the subject, it can only
be determined in certain ways
\enumsentence{*Det sitter den fågel på taket. \\
              **Det sitter fåglarna på taket.}
The distinction lays in the definitness of the noun itself, (not the whole noun phrase..)
it must be indefinite, even though determiners are allowed. Jfr
\enumsentence{Det sitter några fåglar på taket\\
              **Det sitter vardera fåglarna på taket}
%%Quantifiers like \emph{`denna'} or determiners like \emph{`samtliga'} may
%%not be used, but the determiners \emph{`många'}, \emph{`några'} and the quantifier
%%\emph{`ingen'} work well.
%%Semantic difference, atm we allow any NP. \\
The grammar handles this by looking at the determiner used
\verb|FormalSub : SimpleVP -> Det -> NP -> Cl ;|
The function combines a verb phrase a determiner and a noun phrase in to a clause.
\verb|SimpleVP| is a normal VP, but with the restriction that the main verb must be
intransitiv, a \verb|V|, (and passives?). If the  determiner requires a noun in determinened
form, the clause is put to NONEXIST.

'det är kul att..' Näst vanligaste två-ordskombinationen i talad svenska.
%%
%embedded - relative, 'sådan att' not really nice for normal swedish, and no agreement
%between subject 'en katt sådan att det regnar'.
pronouns/Quant. Many, created categories for 'sådan', and 'fler'. Not many in each,
has to decide, one category for each?
Genitive -problem 1 cannot be alone (fixed) 2 some words have id. (no) gen form
           fixing this include restructuring all
Det-utr. not effect for the some Dets, like 'sådan', no difference between utrum and neutrum.
 Can vary in number, since is a NP. Do not want to cut connection to abstract but changing them,
 so added one more. Leads to ambiguities in some cases, but considering that we would want to use
 this for translation, neutr and utr may differ in the other language.

@@

'han såg sin mammas bil', they same could be applied to
APs for example \ref{ex:ReflAP} or Adv for
\enumsentence{Han leker oftare än sin kompis}
 (ComparAdvAdj)
Atm you can't focus this kind of object, then a ClSlash need info about the
subject. FunRP also tricky (and what does it mean, we can say
Han såg sin katt i vilken musen låg), in this case RP needs to give more information.
Kort sagt, the information about the subject must be spread
in different parts of the sentence. 
Could also decide to always produc 'sin' -> jag såg sin bil
to avoid the dependence, but not nice. Or keep a flag in NP saying if it is
in third person, and add nonexist for others. But this way we can also get rid
of ReflVP and the possessive prons \verb|PossPron| should only be used for somebody else:
'jag såg din bil'.
Unfortunaly, have to change ReflVP, this only allows this in object for VPSlash, not FocObj,
not 'alla sina', usw. Want it to handle conjuction as in example \ref{ex:refGen}
etc.
%%

@@

\subsection{Results and evalutaion}
The flat version of Talbanken05 was used for developping the mapping, but
the program works for both versions.
The deeper contains six more tags, but they are not always useful for our       
purpose. Figuer \ref{fig:mappDeepFlat} shows the difference for a conjunction.
% show piece of s5 here!
%\begin{figure}[h]
%\centering
%\subfloat[Talbanken tree]{\label{pic:tbtree}\includegraphics[width=60mm]{Talbankentree.pdf}}
%\subfloat[GF abstract tree]{\label{pic:gftree}\includegraphics[width=60mm]{gfTree.pdf}}
%\caption{Ttrees for the sentence ``Katten på bilen blir större".}
%\label{fig:translationtrees}
%\end{figure}

@@

\chapter{Summary}
\section{Results and discussion}
Results from mapping, can probably be improved, bit by bit, depends much on words.
Valencys would therefor help.  
Saldo, made a big difference to renew the lexicon. Tested what was missing now from talbanken,
results. Mostly compound nouns, which we can't expect in the dictionary.

@@

\section{Future Work}
\label{sec:future}
What is still left to do? Description and direction of future work:
Handling unknown words and , handling unknown grammar, 
how to make the parser robust, elipses and long sentences, names and numbers.
What is still needed in the grammar, how important that is. 
\subsection{Valency}
\label{sec:futureValency}
how to get valency information 'sitter och läser',
is this doable for swedish?
description of extracting this from talbanken/korp
or handling it by the robust parser or the grammar itself (leaveout obj?).
make the parser faster by using smaller lexicon.
Lexicon tool on the fly.
\subsection{Probabilites}
\label{sec:futureProbabilities}
Getting and using probabilities, to disambiguate.
process list from saldo.
Use multiligual treebank?
stop ambiguities by fixing so that Gen of a quant isn't a quant, or better, that all fields
have a genitive field.
More fields in the grammar? Or leave it all to the robust parser

>>>>

\documentclass{report} 
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage[margin=1.5in]{geometry}
\usepackage{lingmacros}
\usepackage{verbatim}
\usepackage{subfig}
% exempel: 2) Inte vet jag
%              not know I
%             I don't know

@@

\subsection{Swedish}
Swedish is a North-Germanic language spoken,
closely related to Norweigan and Danish. The languages share most of their
grammatical structures and are mutual intelligibility. Swedish is also 
an official language in Finland and spoken by approximately 10 million people.
A list of things:
\subsubsection*{Verb second}
Swedish is a verb-second language: the
second constituent of a declarative main clause must consist of a verb.
The normal word order is Subject-Verb-Object, but any syntactic category can be fronted.
The examples \ref{ex:swedish-svo} - \ref{ex:swedish-avso} all have the same propositional
meaning, although the fronted part of sentence \ref{ex:swedish-ovs} are being emphasized.
\enumsentence{
\shortex{4}
{Jag & ser & dig & inte.}
{I &  see & you & not.}
{`I  don't see you'.}} \label{ex:swedish-svo}
\enumsentence{
\shortexnt{4}
{Dig  & ser &  jag & inte.}
{You& I& see & not.}}\label{ex:swedish-ovs} 
\enumsentence{
\shortexnt{4}
{Inte & ser& jag &dig.}
{Not&  see& I&   you.}} \label{ex:swedish-avso}

%\shortexm{3}
%{hej} {hi}{hhe} {bing}{bong}{salt}
%; adverbials, objects
%Alla syntaktiska kategorier kan be fronted utom obetonade satsadverbial (ju,väl).
%The first constituent of the clause is usually made up of the subject,
%although it likewise could consist of adverbial phrases or objects that are
%being empathazied. \\
%\enumsentence{
%\begin{tabular}{lll}
%  Johan & gick & sakta på gatan.\\
%\emph{Johan} & \emph{walked} & \emph{slowely on the street.}\\
%  Sakta & gick & Johan på gatan.\\
%\emph{Slowly} & \emph{walked} & \emph{Johan on the street.}\\
%  På gatan & gick & Johan sakta.\\
%\emph{On the street} & \emph{walked}  & \emph{Johan slowly.}\\
%\end{tabular}}

Inverted word order marks questions
\enumsentence{
\shortex{4}
{Såg & jag & dig & inte?}
{Saw  & I&  you& not?}
{`Didn't I see you?'}}
%\enumsentence{Gick Johan på gatan?\\
%        \emph{Did Johan walk on the street?}}
The word order in subordinate clauses in also slightly different,
%\enumsentence{
%\begin{tabular}{ll}
%{\shortex{7}
%{Jag & såg&  Anna.&  Hon & såg & inte&  mig.} 
%{ & & &  & & &}}
%\shortex{7}
%{Jag såg att Anna \textbf{inte} såg mig.}\\
%\shortex{7}
%{I  saw Anna. She saw not   me.}  & {I   saw that Anna not saw me}. \\
%`I saw Anna. $\;$ She did not see me.'& `I saw that Anna did not see me'.
%}
\enumsentence{Main sentences:\\
{\shortex{7}{Jag & såg & Anna. & Hon & såg & {\bf inte} & mig }
{ I & saw & Anna. & She & saw & {\bf not} & me}
{`I saw Anna. She didn't see me'}}}
\enumsentence{Subordinate sentence:\\
\shortex{7}{Jag & såg & att& Anna& \textbf{inte}& såg& mig.}
            {I &  saw & that& Anna& \textbf{not}& saw& me.}
            {`I saw that Anna didn't see me'}}

\subsubsection*{Passive voice}
Passive voice is often used in Swedish, especially the
\textit{s-passive}, which is formed by adding an \emph{s} at the end of a verb.
\enumsentence{
\shortex{5}{Han & jagade\textbf{s} & av & ett & lejon.}
        {He & hunted+\textbf{s} & by & a & lion.}
        {`He was hunted by a lion.'}}
cf.
\enumsentence{
\shortex{5}{Ett & lejon & jagade & honom.}
        {A &lion &hunted &him.}
        {`A lion was hunting him.'}}

\subsubsection*{Impersonal constructions}
Constructions with 'det är/var' ('it is/was') is very common\cite{H&H309d}:
\enumsentence{
\shortex{6}{Det & var & roligt & att & höra.}
{It &was & nice &to &hear.}
{`I'm glad to hear that.'}}
'Det' (\emph{it}) is also used as formal subject %for intransitive verbs,
with the real subject put in the position of an object.
\enumsentence{
\shortex{6}{Det &står &en &älg &på &fältet.}
{ It & stands &a & moose &on &the field.}
{`There is a moose on the field.'}}

\subsubsection*{Reflexive pronouns}
The Scandinavian language have reflexive pronouns\cite{H&H310}
and special reflexive forms of the possessive pronouns for the 3rd
person\cite{H&H319}. Those are distinct from the normal 3rd person forms.
\enumsentence{
\begin{tabular}{ll}
Han slog sig. & Han såg sitt barn.\\
\vspace{3mm}
\emph{He hit him self.} & \emph{He saw his (own) child.}\\
Han slog honom. & Han såg hans barn.\\
\emph{He hit him (another person).}&\emph{He saw his (another person's) child.}
\end{tabular}}
The 1st and 2nd persons use the normal personal pronoun in object form.

\subsection{Saldo}
A good parser needs a good lexicon. We have used of Saldo\cite{saldo}, a
large electronic lexicon developed and maintained at Gothenburg University. It is
built upon Svenskt Associations Lexicon and contains information about modern
Swedish and the words are equipped with both sematic and a morphological
information. The user can find compounding analyses, examples of usage in corpora,
and graphs of sematically connected words.\\
Example of katt? \\
For our purpose however, only the morphological data was needed.\\
Example of morpho for katt.\\
Saldo provides this in XML format under LPGL. The data can be processed and
translated to GF format as described in 
section \ref{sec:prog.saldo}.
%Other usages?

@@

Being developed with multilinguality in mind, GF makes an important distiction between
\textit{abstract} and \textit{concrete} syntaxes. The abstract is meant to be
shared by many languages, it models the semantics of the domain without blanda in
language specific features such as agreement or word order.

\begin{figure}[h]
\begin{verbatim}
 abstract TestGrammar = {
  cat N ; V ; S ;

  fun 
    Pred : N -> V -> S ;
    cat_N : N ;
    sleep_V : V ;
 }
\end{verbatim}
\caption{A small example of an abstract syntax}
\end{figure}

The example shows an abstract grammar defining three categories, % \verb|Categories|,
one for nouns, one for verbs and one for sentences. An abstract grammar also gives
the types of functions. In this case we have \verb|Pred| which
tells us that by taking a noun and a verb we can form a sentence. At this stage, 
no information of how this is done is given. The grammars also defines two words,
\verb|cat| and \verb|sleep|.

% or how any of the categories should look.
\begin{figure}[h]
\begin{verbatim}
concrete TestGrammarSwe of TestGrammar = {
  lincat N, V, S = Str ;

  lin Pred n v   = n ++ v ;
      cat_N   = "katten"  ;
      sleep_V = "sover"  ;
}
\end{verbatim}
\caption{A Swedish concrete grammar}
\label{fig:gfSweCnc1}
\end{figure}

Figure \ref{fig:gfSweCnc1} shows how the abstract grammar can be implemented
for Swedish. Nouns, verbs and sentences are all defined as strings, \verb|Str|
and \verb|Pred| simply glues the two strings ``katten" and ``sover" together:
\verb|Pred cat sleep = "katten sover"|.\\
Let's look at a more complicated example, where nouns can be used in both
plural and singular. To do this, we add a category \verb|NP| to the
abstract. \verb|N| now means a noun
in any of the number, and we introduce two funciton, \verb|NSg : N -> NP| and
\verb|NPl : N -> Pl| which sets the number.
\begin{figure}[h!]
\begin{verbatim}              
 abstract TestGrammar = {          concrete TestGrammarSwe of TestGrammar = {
  cat N ; NP ; V ; S ;               lincat V, S, NP = Str ;
                                            N = {s : Num => Str} ;
  fun                                lin   
    Pred : NP -> V -> S ;              Pred n v = n ++ v ;
    NSg : N -> NP ;                    NPl n = n.s ! Pl ;
    NPl : N -> NP ;                    NSg n = n.s ! Sg ;
    cat_N : N ;                        cat_N = {s = table {Sg => "katten" ;
    sleep_V : V ;                                          Pl => "katterna"}};
 }                                     sleep_V = "sover"  ;
                                     param Num = Sg | Pl ;
                                     }
\end{verbatim}           
\caption{A modified grammar}
\label{fig:gfTest2}
\end{figure}
Figure \ref{fig:gfTest2} introduces some new concepts: records,tables and parameters.
In the concrete syntax, \verb|N| is defined to be a record consisting of the field
\verb|s|. The type of \verb|s| shows that it is a table, which given a parameter of type \verb|Num| returns
a string. \verb|Num| is defined as a parameter, which can either have value \verb|Sg|
or \verb|Pl|. In \verb|NPl| and \verb|NSg|, \verb|n.s| means that we use the
field \verb|s| of \verb|n|, and then the selection operator \verb|!| is used to
select a branch in the table.

If we want to implement a English version of the grammar, we encounter another
problem: the verb form depends on the noun phrase. The \verb|NP| needs to carry
information about which number it is in, and \verb|Pred| can pass this on to the
verb. Hence, the \verb|V| also needs to be a table, depending on the number.
\begin{figure}[h!]
\begin{verbatim}              
concrete TestGrammarEng of TestGrammar = {
  lincat S = Str ;
         V = {s : Num => Str} ;
         N = {s : Num => Str} ;
         NP = {s : Str ; num = Num} ;
  lin   
    Pred n v = n.s ++ v.s ! n.num ;
    NPl n = {n.s ! Pl ; num = Pl} ;
    NSg n = {n.s ! Sg ; num = Sp} ;
    cat_N = {s = table {Sg => "the cat" ;
                        Pl => "the cats"}};
    sleep_V = {s = table {Sg => "sleeps" ; 
                          Pl => "sleep"}};
  param Num = Sg | Pl ;
  }
\end{verbatim}           
\caption{English implementation}
\label{fig:gfTestEng}
\end{figure}

%maybe continue example with swedish and english, have DetNP : N -> N wich
%gives 'the cat', 'katten'. Thereby introduce tables, parameters, fields.
%Show abstract tree.

We now have two implementations of the abstract. The resulting GF grammar is able to
both parse a string to the abstract tree and to go in the other direction; to produce
a string of natural language given a abstract tree. This step is called linearization.
Translation is a consequence of this, we can parse a Swedish string and then 
linearize the resulting abstract tree to English.

@@

A GF grammar usually covers a very restricted set of language where few ambiguities
occur and much care can be taken to preserving the semantics of the
different languages.
By using a controlled language like this,  good and reliable translation can be
ensured. Some examples of projects using GF are the math .. \cite{olga}, the
Online Syllogism Solver\cite{malin!}.  The framework is also used in the
european project MOLTO for online translation between upto 15 language.\\

\subsubsection{The resource library}
The resource library\cite{gf-resource} is a very useful part of the framework. It is provided in the 
GF package. %, they give a good start for all who want to write an application grammar.
Instead of letting the users of the framework reimplement the morphological and
the fundamental syntactic structures of a language for each application, all
this can be found as libraries. So far there are 20 languages in the library, and
they all share the same abstract syntax.
The resource grammars describe how to decline words and how to construct
phrases and sentences.
There are also a small test lexicon included, containing some hundered of common words.\\

@@

%inte vet jag. foc + v + resten

@@

@@

\subsection{GF for Swedish}
The GF resource grammars contains a fundamental description of Swedish,
covering the morphology, many phrases and clauses.
Covers sentences like
han har inte ätit två gula äpplen idag
word order, agreement, tempus, adverbs, adjectivs...
and relatives
Also info in the Extra module, where more language specific constructions
can be found.
jag ser dig
dig ser jag
dig har jag inte sett
jag har inte sett dig.
how the parse tree is generated, the finite verb, the subject, the rest. usw.
han såg katten han tyckte om
för vem måste jag akta mig
vem måste jag akta mig för

honom ville hon inte tänka på.

@@

For verbs, but also others there is an important distinction for valency.
Either directly to VP if V, or needs to be given arguments. Can then add adverbs.
Maybe talk about ställa.
Particles and prepositions, can't be seen on the trees.
tycka om /tycka om -> om vad tycker du inte. fast bra

@@

%%PASSIV
Is passive also more used in Swedish?
There are two ways of forming passive verb phrases in Swedish: the 
periphrastic passive, formed by using a modale auxiliary verb 
\enumsentence{Bilden blev tagen av  av rymdsonden Galileo}
\enumsentence{The picture was taken by spacecraft Galileo.}
or by adding an \emph{s} to
the verb:
\enumsentence{Bilden togs av rymdsonden Galileo}
\enumsentence{The picture [took+s] by spacecraft Galileo.}
This second variant, the s-passive, is commonly used in Swedish, some studies
suggest that it is used in more than 80\% of the times \cite{laanemets}.
For the other scandinavian languages, the s-passive can not be used in all tenses for all words.
The periphrastic passive is preferred, especially in spoken langauge.
\enumsentence{Bildet ble tatt av romsonden (NO)}
The resource grammar for Scandinavian therefore allowed/expressed constructions using auxiliary verb. \\
\verb|PassV2 : V2  -> VP | \\
\verb|         äta -> blev äten| \\
This rules allows two-place verbs to be used in passive by using \emph{bli} (\emph{become}), and thereby
turned into comlpete verb phrases; they no longer need an object.
The s-passive has now been added as the standard case for the Swedish GF
grammar. Periphrastics passive is still allowed, but as an alternative rather
than the default.
The grammar also allows using three-place verbs \verb|V3| , as \emph{give} to
form passives:
\enumsentence{Betyget 3 ges till många elever}
\label{ex:passV3}
by letting the first object be `removed' and given by the subject. 
Sentences where the second object is put first in the sentence, like 
\enumsentence{Till många elever ges betyget 3}
is a variant of example \ref{ex:passV3}, \emph{`många elever'} is still the object but put in focus. The grammar
parse this by applying \verb|FocObj|.
%%

@@

Since those constructions are inspired by other languages, some of them may sound quite
non-natural in Swedish, such as the \verb|RelS|. This combines a sentence \emph{`Hon sover'}
and a relative sentence \emph{`som/vilket är bra'} which wrongly expressed 
\emph{`She sleeps, which is good'} as
\emph{`Hon sover som är bra'}. The error comes from the fact that \emph{`som'} is
embedded in the relative clause. To fix this, an extra parameter was needed to tell whether
\emph{`som'} or \emph{`vilket'} should be used. \\
Another complication stems from \verb|RelCl| which express the english \emph{`such that'}.
The Swedish version \emph{`sådan att'} sounds stelbent, when used outside of logic books.
\enumsentence{Jag vill ha en katt sådan att den inte fäller.} 
An more natural sounding alternative would be
\enumsentence{Jag vill ha en sådan katt som inte fäller.}
and hopefully this will be implemented soon, otherwise: why is it hard and how we can say
'en katt sådan att det regnar'. It is relative but has on empty field, still congruation is
needed.
du är här, vilket jag vet.
i de äpplen/det äpple du äter bor en mask/ som du ser.\\

\subsection{Presentering, Impersonal constantiation(?), Topicalisering}
Formal subjects \cite{SAG-19,H&H-309d} is often used in Swedish
\enumsentence{Det sitter en fågel på taket}
\emph{`There is a bird sitting on the roof'}
or, literally
\emph{`It sits a bird on the roof'}.\\
\emph{`Det'} has the position of the subject, and the real subject, 
\emph{`en fågel'} the one of an object.
Transitive verbs may not be used like this
\enumsentence{*Det äter en fågel på taket.}
unless their in passive form
\enumsentence{Det äts en fågel på taket.}
%det finns/saknas/fattas
There are also restrictions on the nounphrase of the subject, it can only
be determined in certain ways
\enumsentence{*Det sitter den fågel på taket. \\
              **Det sitter fåglarna på taket.}
The distinction lays in the definitness of the noun itself, (not the whole noun phrase..)
it must be indefinite, even though determiners are allowed. Jfr
\enumsentence{Det sitter några fåglar på taket\\
              **Det sitter vardera fåglarna på taket}
%%Quantifiers like \emph{`denna'} or determiners like \emph{`samtliga'} may
%%not be used, but the determiners \emph{`många'}, \emph{`några'} and the quantifier
%%\emph{`ingen'} work well.
%%Semantic difference, atm we allow any NP. \\
The grammar handles this by looking at the determiner used
\verb|FormalSub : SimpleVP -> Det -> NP -> Cl ;|
The function combines a verb phrase a determiner and a noun phrase in to a clause.
\verb|SimpleVP| is a normal VP, but with the restriction that the main verb must be
intransitiv, a \verb|V|, (and passives?). If the  determiner requires a noun in determinened
form, the clause is put to NONEXIST.

finnas: här får man ha definit: det finns de som inte vill. Samma RelCl som
jag just la till, de äppplen. (de + relsats)

'det är kul att..' Näst vanligaste två-ordskombinationen i talad svenska.
%%
%embedded - relative, 'sådan att' not really nice for normal swedish, and no agreement
%between subject 'en katt sådan att det regnar'. Kan säga 'en sån katt som sover, med 'sådan' + Rel
%har lagt till Det äpple + Rel (RelCNNP). *Det äpple är rött. men Det äpple jag äter är rött.
%

pronouns/Quant. Many, created categories for 'sådan', and 'fler'. Not many in each,
has to decide, one category for each?
Genitive -problem 1 cannot be alone (fixed) 2 some words have id. (no) gen form
           fixing this include restructuring all
Det-utr. not effect for the some Dets, like 'sådan', no difference between utrum and neutrum.
 Can vary in number, since is a NP. Do not want to cut connection to abstract but changing them,
 so added one more. Leads to ambiguities in some cases, but considering that we would want to use
 this for translation, neutr and utr may differ in the other language.

@@

'han såg sin mammas bil', they same could be applied to
APs for example \ref{ex:ReflAP} or Adv for
\enumsentence{Han leker oftare än sin kompis}
 (ComparAdvAdj)
Atm you can't focus this kind of object, then a ClSlash need info about the
subject. FunRP also tricky (and what does it mean, we can say
Han såg sin katt i vilken musen låg), in this case RP needs to give more information.
Kort sagt, the information about the subject must be spread
in different parts of the sentence. 
Could also decide to always produc 'sin' -> jag såg sin bil
to avoid the dependence, but not nice. Or keep a flag in NP saying if it is
in third person, and add nonexist for others. But this way we can also get rid
of ReflVP and the possessive prons \verb|PossPron| should only be used for somebody else:
'jag såg din bil'.
Unfortunaly, have to change ReflVP, this only allows this in object for VPSlash, not FocObj,
not 'alla sina', usw. Want it to handle conjuction as in example \ref{ex:refGen}
etc.
varandra, works the same but has no case for Sg.
Johan vill att jag rakar sig* -> honom  
%%

@@

\subsection{Results and evalutaion}
The flat version of Talbanken05 was used for developping the mapping, but
the program works for both versions.
The deeper contains six more tags, but they are not always useful for our       
purpose. Figure \ref{fig:mappDeepFlat} shows the difference for a conjunction.
% show piece of s5 here!
\begin{figure}[h]
\centering
\subfloat[Deep]{\label{pic:mappDeep}\includegraphics[width=20mm]{nofile.png}}
\subfloat[Flat]{\label{pic:mappFlat}\includegraphics[width=20mm]{nofile.png}}
\caption{...}
\label{fig:mappDeepFlat}
\end{figure}

@@

\chapter{Summary}
\section{Results and discussion}
Results from mapping, can probably be improved, bit by bit, depends much on words.
Valencys would therefor help.  
Saldo, made a big difference to renew the lexicon. Tested what was missing now from talbanken,
results. Mostly compound nouns, which we can't expect in the dictionary.
Is too slow, can't be used for parsing right away

@@

\section{Future Work}
\label{sec:future}
What is still left to do? Description and direction of future work:
Handling unknown words and , handling unknown grammar, 
how to make the parser robust, elipses and long sentences, names and numbers.
What is still needed in the grammar, how important that is. 
\subsection{Lexicon/Parallell parsing}
Speed up.
\subsection{Valency}
\label{sec:futureValency}
how to get valency information 'sitter och läser',
is this doable for swedish?
description of extracting this from talbanken/korp
or handling it by the robust parser or the grammar itself (leaveout obj?).
make the parser faster by using smaller lexicon.
Lexicon tool on the fly.
\subsection{Probabilites}
\label{sec:futureProbabilities}
Getting and using probabilities, to disambiguate.
process list from saldo.
Use multiligual treebank?
stop ambiguities by fixing so that Gen of a quant isn't a quant, or better, that all fields
have a genitive field.
More fields in the grammar? Or leave it all to the robust parser
